Modeling Predator Preferences

### 2014-11-29 paper/data.tex
edit and typo check

We analyze a dataset that was collected to investigate the eating preferences of the wolf spider, genus \textit{Schizocosa}, towards the prey orders \textit{Diptera} and \textit{Collembola}.  Predators were collected in traps within the deciduous Berea College Forest in Madison County, Kentucky, USA and had their gut-contents analyzed to determine whether or not the spiders ate any \textit{Diptera} or \textit{Collembola} within each time period.  The left panel of figure~\ref{fig:data} plots the percentage of spiders that had the prey in their guts across time.  Prey were also collected in traps from the same area.  These data were collected for each month from October $2011$ to March $2013$.  On average, $69$ spiders, and $111$ and $297$ \textit{Diptera} and \textit{Collembola}, respectively, were caught in each time period.  The range of the sample sizes, across all $18$ months is, $11$ to $181$ for caught spiders, from $7$ to $322$ for trapped \textit{Diptera}, and from $101$ to $755$ for trapped \textit{Collembola}.  The right panel of figure~\ref{fig:data} plots the total number of each order that was caught during each time period.

@@

These data provide an example of our hierarchy of hypotheses.  First, tested model $c_{st} = c$ against $c_{st} = c_s$, to determine whether or not the wolf spider has different preferences for the two orders \textit{Diptera}  and \textit{Collembola}.  With, one degree of freedom, this likelihood ratio test indicated, $p-value < 0.0001$,  that two parameters, one for each order, fits these data better than one parameter for both.  Similarly, we tested whether or not there was a significant effect across time by testing model $c_{st} = c$ against $c_{st} = c_t$.  Here, the likelihood ratio test, with $17$ degrees of freedom, implies that the wolf spiders of the Berea College Forest eat these prey orders at different rates across the months of the year, $p-value < 0.0001$.  In fact, we find that the most parameter rich model, $\lambda_{st} = c_{st} \gamma_{st}$ fits these data better than is expected by chance, $p-value < 0.0001$.  Model $c_{st}$ estimates $72$ parameters in total; since, in this case, there are two prey of interest and $18$ time periods, it takes $36$ parameters to estimate each $c_{st}$ and $\gamma_{st}$.  Figure~\ref{fig:cst} plots the point estimates and confidence intervals of $c_{st}$, for both prey across all time periods.

\begin{figure}
  \centering
  \includegraphics[scale=0.5]{cst}
  \caption{One plot for each prey order displays the point estimates and confidence intervals as estimated from the model $c_{st}$.}
  \label{fig:cst}
\end{figure}

>>>>

We analyzed a dataset that was collected to investigate the eating preferences of the wolf spider, genus \textit{Schizocosa}, towards the prey orders \textit{Diptera} and \textit{Collembola}.  Predators were collected in traps within the deciduous Berea College Forest in Madison County, Kentucky, USA and had their gut-contents analyzed to determine whether or not the spiders ate any \textit{Diptera} or \textit{Collembola} within each time period.  The left panel of figure~\ref{fig:data} plots the percentage of spiders that had the prey in their guts across time.  Prey were also collected in traps from the same area.  These data were collected for each month from October $2011$ to March $2013$.  On average, $69$ spiders, and $111$ and $297$ \textit{Diptera} and \textit{Collembola}, respectively, were caught in each time period.  The range of the sample sizes, across all $18$ months is, $11$ to $181$ for caught spiders, from $7$ to $322$ for trapped \textit{Diptera}, and from $101$ to $755$ for trapped \textit{Collembola}.  The right panel of figure~\ref{fig:data} plots the total number of each order that was caught during each time period.

@@

These data provide an example of our hierarchy of hypotheses.  First, we tested model $c_{st} = c$ against $c_{st} = c_s$, to determine whether or not the wolf spider has different preferences for the two orders \textit{Diptera}  and \textit{Collembola}.  With, one degree of freedom, this likelihood ratio test indicated, $p-value < 0.0001$,  that two parameters, one for each order, fits these data better than one parameter for both.  Similarly, we tested whether or not there was a significant effect across time by testing model $c_{st} = c$ against $c_{st} = c_t$.  Here, the likelihood ratio test, with $17$ degrees of freedom, implies that the wolf spiders of the Berea College Forest eat these prey orders at different rates across the months of the year, $p-value < 0.0001$.  In fact, we find that the most parameter rich model, $\lambda_{st} = c_{st} \gamma_{st}$ fits these data better than is expected by chance, $p-value < 0.0001$.  Model $c_{st}$ estimates $72$ parameters in total; since, in this case, there are two prey of interest and $18$ time periods, it takes $36$ parameters to estimate each $c_{st}$ and $\gamma_{st}$.  Figure~\ref{fig:cst} plots the point estimates and $95\%$ confidence intervals of $c_{st}$, for both prey across all time periods.

\begin{figure}
  \centering
  \includegraphics[scale=0.5]{cst}
  \caption{One plot for each prey order displays the point estimates and $95\%$ confidence intervals as estimated from the model $c_{st}$.}
  \label{fig:cst}
\end{figure}

### 2014-11-29 paper/discussion.tex
edit and typo check

The model developed here allows practicioners to determine predators' preferences by testing simultaneously across an array of multiple prey species and time.  This is acheived via a simple, but statistically powerful, likelihood ratio test.  Further testing of the ratio of rates for which predators eat to encounter prey species allows researchers to make specific conclusions about predators' preferences.  For instance, rates acros time can be estimated to make statements about seasonal effects on a predator's eating habits, or relative rates across species groups allows for statements about the preferences for different species.

When counts of predators' gut contents are not fully observed, and instead only a binary response indicating the existence of the prey species in the gut is observed, we are able to treat the counts as missing data.  By modelling all of the observed data, both the binary responses and the number of prey species caught, and the missing count data, we are able to use the EM algorithm to extract as much information from the data as possible.  Though this is nice in theory, in practice the success of this modification to our original model is limited by the magnitude of the unknown parameters $\lambda_{st}$.

Further developments of our model could be beneficial.  Taking into account other environmental variables that might effect a predators' eating habits, such as rain or temperature, say, might be advantageous.

>>>>

The model developed here allows practitioners to determine predators' preferences by testing simultaneously across an array of multiple prey species and time.  This is achieved via a simple, but statistically powerful, likelihood ratio test.  Further testing of the ratio of rates for which predators eat to encounter prey species allows researchers to make specific conclusions about predators' preferences.  For instance, rates across time can be estimated to make statements about seasonal effects on a predator's eating habits, or relative rates across species groups allows for statements about the relative preferences for different species.

When counts of predators' gut contents are not fully observed, and instead only a binary response indicating the existence of the prey species in the gut is observed, we are able to treat the counts as missing data.  By modeling all of the observed data, both the binary responses and the number of prey species caught, and the missing count data, we are able to use the EM algorithm to extract as much information from the data as possible.  Though this is nice in theory, in practice the success of this modification to our original model is limited by the magnitude of the unknown parameters $\lambda_{st}$.

Further developments of our model could be beneficial.  Taking into account other environmental variables that might effect a predator's eating habits, such as rain or temperature, say, might be advantageous.

### 2014-11-29 paper/introduction.tex
edit and typo check

\section*{Introduction}
\label{sec:intro}
%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End:

>>>>

\section{Introduction}
\label{sec:intro}

Modeling a predator's food selection was given much attention for a short period of time; see \citet{Strauss:1979} and the references within.  Many, if not all, of the indices developed, though intuitive, focused on a snapshot in time and on only one prey species, and rarely obtained practical asymptotic properties.  This restricts the interested practitioner to the most computationally manageable of techniques, and has completely ignored simultaneous testing across an array of both species and time.

The model presented here maintains tractable asymptotic properties while being general enough to take into account multiple species and time points.  By modeling both time and any number of prey species, we are able to see a more detailed analysis of the predator's eating preferences.  The simplicity of the model allows us to consider predators for which exact tallies of the number of each prey species eaten within any given time period is not observed.  Instead, we rely on the researcher being able to DNA sequence the contents of the predator's gut and make a simple binary conclusion: this predator ate some of that prey species during this time period, or did not.  Despite the data's lost information, under certain situations our model is able to accurately estimate the parameters of interest.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2014-11-29 paper/methods.tex
create option to store random seed in testPref
edit and typo check

The use of Poisson distributions make the following implicit assumptions: $1)$ traps independently catch the prey species of interest, $2)$ predators eat prey species independently, $3)$ predators eat indepedent of each other.

We denote the number of predators and the numer of prey species caught, in each time period $t$, by $J_t$ and $I_t$, respectively.  Let $X_{jst} \iid \mathcal{P}(\lambda_{st})$ represent the number of prey species $s$ that predator $j$ ate during occurrence $t$, where $j \in \{1, \ldots, J_t\}$.  Let $Y_{ist} \iid \mathcal{P}(\gamma_{st})$ represent the number of prey species $s$ found in trap $i$ during occurrence $t$, $i \in \{1, \ldots, I_t\}$.  Formal statistical statements about the relative magnitudes of the parameters $\boldsymbol{\lambda}$ and $\boldsymbol{\gamma}$ offer insights to the relative rates at which predators eat particular prey species.

@@

\begin{enumerate}
\item $c_{st} = 1, s = 1, \ldots, S; t = 1, \ldots, T$
\item $c_{st} = c, s = 1, \ldots, S; t = 1, \ldots, T$
\item $c_{st} = c_s, s = 1, \ldots, S$
\item $c_{st} = c_t, t = 1, \ldots, T$
\item $c_{st} = c_{st}, s = 1, \ldots, S; t = 1, \ldots, T$
\end{enumerate}

The first hypothesis states that predators and traps sample all prey species at the same rate.  One imagines this is the case if the predator simply eats that which comes within its reach, thus suggesting a diet indifferene.  The second says that predators sample prey proportionally across all time periods.  The third hypothesis says that predators sample different prey species at different rates, but each rate is steady across time.  This implies that the predator expresses preferences for one prey species over another, but is unresponsive to changes due to time.  Conversely, the forth hypothesis implies that each prey species is sampled similarly within each time period, while the rates across time are allowed to change.  The fifth assumes a predator's selection varies by both time and prey species.  This would make sense if environmental variables, say weather, or prey availability, and taste were affecting predators' selection strategies.

@@

\subsection{Fully Observed Count Data}

@@

\subsection{Unobserved Counts}

@@

Many times it may not be possible to count the number of individuals of each prey species that are in a predators gut. Instead, it may only be possible to detect whether or not a predator consumed the prey species during a given period. This is the case in the study of ? spiders which relied on DNA sequencing to identify the prey species in a spider's gut contents. Despite this, we can still make inference about the predators preferences for the different prey species by using the EM algorithm to compute maximum likelihood estimates.

We denote the binary random variable indicating that the $j^{th}$ predator did in fact eat at least one individual of prey species $s$ in time period $t$ by $Z_{jst} = 1(X_{jst} > 0)$.  The variables are independent Bernoulli observations with success probability $p_{st} = P(Z_{jst}=1)= 1-\exp\{-\lambda_{st}\}$.  Despite not observing $X_{jst}$, we can compute maximum likelihood estimates of the parameters $\boldsymbol{\lambda}, \boldsymbol{\gamma}$ through the EM algorithm using the complete data log-likelihood
\[
l_{comp}(\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \log f_{X,Y,Z}(\boldsymbol x, \boldsymbol y, \boldsymbol z|\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \sum_{s=1}^{S} \sum_{t=1}^T [\sum_{j=1}^{J_t} \log f_{X,Z}(x_{jst},z_{jst}|\boldsymbol{\lambda}) + \sum_{i=1}^{I_t}\log f_Y(y_{jst}|\boldsymbol{\gamma}).
\]

The density of $Y_{jst}$ is exactly as given above and so we focus on deriving the joint density of $X_{jst}$ and $Z_{jst}$. With the distribution of $Z_{jst}$ given above, we can compute $f_{X,Z}(x_{jst},z_{jst}|\boldsymbol{\lambda})$ by noting that $X_{jst}=0$ with probability 1 if $Z_{jst}=0$ and that $X_{jst}|Z_{jst}=0$ has a truncated Poisson distribution with density
\begin{equation*}
  f_{X|Y,Z,\boldsymbol{\lambda},\boldsymbol{\gamma}}(x_{jst}|z_{jst}) =
  \frac{\exp{\{-\lambda_{st}\}} \lambda_{st}^{x_{jst}}}{(1 - \exp{\{-\lambda_{st}\}}) x_{jst}!}1(x_{jst} > 0) \quad \text{ where } 
\end{equation*}
and expected value
\[
\E_{X|Y,Z}X_{jst} = \frac{\lambda_{st} \exp{\{\lambda_{st} \}}}{\exp{\{ \lambda_{st} \}} - 1}.
\]
\noindent The joint density of $X_{jst}, Z_{jst}$ is then 
\begin{equation*}
    f_{X,Z|\boldsymbol{\lambda}}(x_{jst},z_{jst}) = \left\{
    \begin{array}{lr}
      \exp{\{ -\lambda_{st} \}}, & x_{jst}=0 \mbox{ and } Z_{jst} = 0 \\
      \frac{\exp{\{-\lambda_{st} \}} \lambda_{st}^{x_{jst}}}{x_{jst}!}, & x_{jst} > 0 \mbox{ and } Z_{jst} = 1\\
      0 & \mbox{otherwise}
    \end{array}
  \right..
\end{equation*}

@@

The EM algorithm works by iterating two steps, the E-step and M-step, until the optimum is reached \citep{Dempster:1977,McLachlan:2007}.  Let $k$ index the iterations in the EM algorithm so that $\boldsymbol \lambda^{(k)}$ and $\boldsymbol \gamma^{(k)}$ denote the estimates computed on the $k^{th}$ M-step. The E-step consists of computing the expectation of $l_{comp}$ with respect to the conditional distribution of $X$ given the current estimates of the parameters
\[
Q^{(k)}(\boldsymbol{\lambda},\boldsymbol{\gamma}) = \E_{X|Y,Z,\boldsymbol{\lambda}^{(k)}} l_{comp}
\]
in order to remove the unobserved data. The M-step then involves maximizing $Q = \E l_{comp}$ with respect to the parameters in the model to obtain updated estimates of the parameters,
\[
(\boldsymbol{\lambda}^{(k+1)}, \boldsymbol{\gamma}^{(k+1)}) = \argmax_{(\boldsymbol{\lambda}, \boldsymbol{\gamma})} Q^{(k)}(\boldsymbol{\lambda}, \boldsymbol{\gamma}).
\]
These steps are then alternated until a convergence criterion monitoring subsequent differences in the parameter estimates/likelihood values is met.

The calculation of $Q^{(k)}(\boldsymbol{\lambda}, \boldsymbol{\gamma})$ is not difficult and is given by:
\begin{align}
  \label{eq:estep}
  \begin{split}
  Q^{(k)}(\boldsymbol{\lambda}, \boldsymbol{\gamma})
  & = \E \log{f_{X,Z|\boldsymbol{\lambda}}(X_{jst},z_{jst})} + \log{f_{Y|\boldsymbol{\gamma}}(y_{ist})} \\
  & = \sum_{s=1}^S \sum_{t=1}^T \sum_{j=1}^{J_t} \E \log{f_{X,Z|\boldsymbol{\lambda}}(X_{jst},z_{jst})}
  + \sum_{s=1}^S \sum_{t=1}^T \sum_{i=1}^{I_t} \log{f_{Y|\boldsymbol{\gamma}}(y)} \\
  & \propto \sum_{s,t,j} \left( - \lambda_{st} 
    + z_{jst} \E X_{jst} \log{\lambda_{st}}  \right) + \sum_{s,t} \left( -I_t \gamma_{st} + Y_{\cdot st} \log{I_t \gamma_{st}} \right) \\
  & \propto \sum_{s,t} \left( -J_t \lambda_{st} + z_{\cdot st} \E (X_{jst}|\lambda_{st}^{(k)},\gamma_{st}^{(k)}) \log{\lambda_{st}} \right) + \sum_{s,t} \left( -I_t \gamma_{st} + Y_{\cdot st} \log{I_t \gamma_{st}} \right).
\end{split}
\end{align}
\noindent No analytic solution to the M-step exists, however, so we chose to maximize $Q$ with coordinate descent \citep{Luo:1992}.  In fact, as we only need find parameters that increase the value of $Q$ on each iteration, we forgo fully iterating to find the maximum and instead perform just one step uphill within each EM iteration.  Since $Q^{(k)}$ is concave and smooth in the parameters $\boldsymbol{\lambda}, \boldsymbol{\gamma}$, we are able to use the convergence of parameter estimates, $||(\boldsymbol{\lambda}^{(k)}, \boldsymbol{\gamma}^{(k)}) - (\boldsymbol{\lambda}^{(k+1)}, \boldsymbol{\gamma}^{(k+1)})||_{\infty} < \tau $, for some $\tau>0$, as our stopping criterion.

As we show in our simulation study, this generalized EM algorithm accurately estimates the parameters when values of $\lambda_{st}$ are relatively small, such that zeros are prevalent in the data $Z_{jst}$.  In this case, not too much information is lost since estimation of $\E Z_{jst}$ can be estimated well by the proportion of observed zeros.  On the other hand, if the predator consistently eats a given prey species, few to no zeros will show up in the observed data and $\E Z_{jst}$ is estimated to be nearly $1$.  The loss of information is best seen by attempting to solve for $\lambda_{st}$ in the equation $\E Z_{jst} = 1 = 1 - \exp\{-\lambda_{st}\}$.  As the proportion of ones in the observed data increase, we expect $\lambda_{st}$ to grow exponentially large.  When no zeros are present in the data, so that where only ones are observed, the likelihood can be made arbitrarily large by sending the parameter off to infinity.

@@

The likelihood ratio test (LRT) statistic is

@@

\subsection{Testing $c_{st}$}

After determining which model best fits the data, more detail can be extracted through a hypothesis test of the elements of $c_{st}$, or in vector notation as $\mathbf{c} \in \mathbb{R}^{S\cdot T}$.  Let the elements of $\hat{\mathbf{c}}$ be the maximum likelihood estimates, $\hat{c}_{st}$, as found via the framework above.  Since $\hat{\mathbf{c}}$ is asymptotically normally distributed, any linear combination of the elements is also asymptotically normally distributed.  For instance, let $a$ be a vector of the same dimension of $\hat{\mathbf{c}}$.  Then $a^t\hat{\mathbf{c}}$ is asymptotically distrbuted as $\mathcal{N}(a^t\mathbf{c}, a^t\Sigma a)$, where $\Sigma$ is the covariance matrix of the asymptotic distrbution of $\hat{\mathbf{c}}$.

Suppose, for example, that the hypothesis $c_s$ is determined to best fit the data with $s$ ranging $s = 1, 2, 3$.  We can test to see whether or not two species are statistically equally preferred under the null hypothesis $c_{1} = c_{2}$.  This hypothesis is alternatively written in vector notation as $a^t\mathbf{c} = 0$, where $a = (1, -1, 0)^t$.  Tests of the following form $H_0: a^t\mathbf{c} = \mu$ against any alternative of interest are then approximate $Z$-tests.  Confidence intervals of any size are similarly, readily obtained.

>>>>

The use of Poisson distributions make the following implicit assumptions: $1)$ traps independently catch the prey species of interest, $2)$ predators eat prey species independently, $3)$ predators eat independent of each other.

We denote the number of predators and the number of prey species caught, in each time period $t$, by $J_t$ and $I_t$, respectively.  Let $X_{jst} \iid \mathcal{P}(\lambda_{st})$ represent the number of prey species $s$ that predator $j$ ate during occurrence $t$, where $j \in \{1, \ldots, J_t\}$.  Let $Y_{ist} \iid \mathcal{P}(\gamma_{st})$ represent the number of prey species $s$ found in trap $i$ during occurrence $t$, $i \in \{1, \ldots, I_t\}$.  Formal statistical statements about the relative magnitudes of the parameters $\boldsymbol{\lambda}$ and $\boldsymbol{\gamma}$ offer insights to the relative rates at which predators eat particular prey species.

@@

\begin{enumerate}
\item $c_{st} = 1$
\item $c_{st} = c$
\item $c_{st} = c_s$
\item $c_{st} = c_t$
\item $c_{st} = c_{st}$
\end{enumerate}

The first hypothesis states that predators and traps sample all prey species at the same rate.  One imagines this is the case if the predator simply eats that which comes within its reach, thus suggesting a diet indifference.  The second says that predators sample prey proportionally across all time periods.  The third hypothesis says that predators sample different prey species at different rates, but each rate is steady across time.  This implies that the predator expresses preferences for one prey species over another, but is unresponsive to changes due to time.  Conversely, the forth hypothesis implies that each prey species is sampled similarly within each time period, while the rates across time are allowed to change.  The fifth assumes a predator's selection varies by both time and prey species.  This would make sense if environmental variables, say weather, or prey availability, and taste were affecting predators' selection strategies.

@@

\subsection{Fully Observed Count Data}
\label{sec:count}

@@

\subsection{Unobserved Counts}
\label{sec:noncount}

@@

Many times it may not be possible to count the number of individuals of each prey species that are in a predator's gut.  Instead, it may only be possible to detect whether or not a predator consumed the prey species during a given period.  In this case we can still make inference about the predators' preferences for the different prey species by using the EM algorithm to compute maximum likelihood estimates.

We denote the binary random variable indicating that the $j^{th}$ predator did in fact eat at least one individual of prey species $s$ in time period $t$ by $Z_{jst} = 1(X_{jst} > 0)$.  The variables are independent Bernoulli observations with success probability $p_{st} = P(Z_{jst}=1)= 1-\exp\{-\lambda_{st}\}$.  Despite not observing $X_{jst}$, we can compute maximum likelihood estimates of the parameters $\boldsymbol{\lambda}, \boldsymbol{\gamma}$ through the EM algorithm using the complete data log-likelihood
\[
l_{comp}(\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \log f_{X,Y,Z}(\boldsymbol x, \boldsymbol y, \boldsymbol z|\boldsymbol{\lambda}, \boldsymbol{\gamma}) = \sum_{s=1}^{S} \sum_{t=1}^T \left[ \sum_{j=1}^{J_t} \log f_{X,Z}(x_{jst},z_{jst}|\boldsymbol{\lambda}) + \sum_{i=1}^{I_t}\log f_Y(y_{jst}|\boldsymbol{\gamma}) \right].
\]

The density of $Y_{jst}$ is exactly as in section~\ref{sec:count} and so we focus on deriving the joint density of $X_{jst}$ and $Z_{jst}$.  With the distribution of $Z_{jst}$ given above, we can compute $f_{X,Z}(x_{jst},z_{jst}|\boldsymbol{\lambda})$ by noting that $X_{jst}=0$ with probability 1 if $Z_{jst}=0$, and that $[X_{jst}|Z_{jst}=0]$ has a truncated Poisson distribution with density
\[
  f_{X|Y,Z,\boldsymbol{\lambda},\boldsymbol{\gamma}}(x_{jst}|z_{jst}) =
  \frac{\exp{\{-\lambda_{st}\}} \lambda_{st}^{x_{jst}}}{(1 - \exp{\{-\lambda_{st}\}}) x_{jst}!}1(x_{jst} > 0)
\]
and expected value
\[
\E_{X|Y,Z}X_{jst} = \frac{\lambda_{st} \exp{\{\lambda_{st} \}}}{\exp{\{ \lambda_{st} \}} - 1}.
\]
\noindent The joint density of $X_{jst}, Z_{jst}$ is then 
\begin{equation*}
    f_{X,Z|\boldsymbol{\lambda}}(x_{jst},z_{jst}) = \left\{
    \begin{array}{lr}
      \exp{\{ -\lambda_{st} \}}, & x_{jst}=0 \mbox{ and } z_{jst} = 0 \\
      \frac{\exp{\{-\lambda_{st} \}} \lambda_{st}^{x_{jst}}}{x_{jst}!}, & x_{jst} > 0 \mbox{ and } z_{jst} = 1 \\
      0 & \mbox{otherwise}
    \end{array}
  \right..
\end{equation*}

@@

The EM algorithm works by iterating two steps, the E-step and M-step, until the optimum is reached \citep{Dempster:1977,McLachlan:2007}.  Let $k$ index the iterations in the EM algorithm so that $\boldsymbol \lambda^{(k)}$ and $\boldsymbol \gamma^{(k)}$ denote the estimates computed on the $k^{th}$ M-step. The E-step consists of computing the expectation of $l_{comp}$ with respect to the conditional distribution of $X$ given the current estimates of the parameters
\[
Q^{(k)}(\boldsymbol{\lambda},\boldsymbol{\gamma}) = \E_{X|Y,Z,\boldsymbol{\lambda}^{(k)}} l_{comp}
\]
in order to remove the unobserved data. The M-step then involves maximizing $Q = \E l_{comp}$ with respect to the parameters in the model to obtain updated estimates of the parameters,
\[
(\boldsymbol{\lambda}^{(k+1)}, \boldsymbol{\gamma}^{(k+1)}) = \argmax_{(\boldsymbol{\lambda}, \boldsymbol{\gamma})} Q^{(k)}(\boldsymbol{\lambda}, \boldsymbol{\gamma}).
\]
These steps are then alternated until a convergence criterion monitoring subsequent differences in the parameter estimates/likelihood is met.

The calculation of $Q^{(k)}(\boldsymbol{\lambda}, \boldsymbol{\gamma})$ is not difficult and is given by:
\begin{align}
  \label{eq:estep}
  \begin{split}
  Q^{(k)}(\boldsymbol{\lambda}, \boldsymbol{\gamma})
  & = \E \log{f_{X,Z|\boldsymbol{\lambda}}(X_{jst},z_{jst})} + \log{f_{Y|\boldsymbol{\gamma}}(y_{ist})} \\
  & = \sum_{s=1}^S \sum_{t=1}^T \sum_{j=1}^{J_t} \E \log{f_{X,Z|\boldsymbol{\lambda}}(X_{jst},z_{jst})}
  + \sum_{s=1}^S \sum_{t=1}^T \sum_{i=1}^{I_t} \log{f_{Y|\boldsymbol{\gamma}}(y)} \\
  & \propto \sum_{s,t,j} \left( - \lambda_{st} 
    + z_{jst} \log{\lambda_{st}} \E X_{jst} \right) + \sum_{s,t} \left( -I_t \gamma_{st} + Y_{\cdot st} \log{I_t \gamma_{st}} \right) \\
  & \propto \sum_{s,t} \left( -J_t \lambda_{st} + z_{\cdot st} \log{\lambda_{st}} \E (X_{jst}|\lambda_{st}^{(k)},\gamma_{st}^{(k)}) \right) + \sum_{s,t} \left( -I_t \gamma_{st} + Y_{\cdot st} \log{I_t \gamma_{st}} \right).
\end{split}
\end{align}
\noindent No analytic solution to the M-step exists, however, so we chose to maximize $Q$ with coordinate descent \citep{Luo:1992}.  In fact, as we only need find parameters that increase the value of $Q$ on each iteration, we forgo fully iterating to find the maximum and instead perform just one step uphill within each EM iteration.  Since $Q^{(k)}$ is concave and smooth in the parameters $\boldsymbol{\lambda}, \boldsymbol{\gamma}$, we are able to use the convergence of parameter estimates, $||(\boldsymbol{\lambda}^{(k)}, \boldsymbol{\gamma}^{(k)}) - (\boldsymbol{\lambda}^{(k+1)}, \boldsymbol{\gamma}^{(k+1)})||_{\infty} < \tau $, for some $\tau>0$, as our stopping criterion.

As we show in our simulation study, this generalized EM algorithm accurately estimates the parameters when values of $\lambda_{st}$ are relatively small, such that zeros are prevalent in the data $Z_{jst}$.  In this case, not too much information is lost since estimation of $\E Z_{jst}$ can be estimated well by the proportion of observed zeros.  On the other hand, if the predator consistently eats a given prey species, few to no zeros will show up in the observed data and $\E Z_{jst}$ is estimated to be nearly $1$.  The loss of information is best seen by attempting to solve for $\lambda_{st}$ in the equation $1 = \E Z_{jst} 1 - \exp\{-\lambda_{st}\}$.  As the proportion of ones in the observed data increases, we expect $\lambda_{st}$ to grow exponentially large.  When no zeros are present in the data, so that where only ones are observed, the likelihood can be made arbitrarily large by sending the parameter off to infinity.

@@

The likelihood ratio test statistic is

@@

\subsection{Linear Transformations of $c_{st}$}

After determining which model best fits the data, more detail can be extracted through a hypothesis test of the elements of $c_{st}$, or in vector notation as $\mathbf{c} \in \mathbb{R}^{S\cdot T}$.  Let the elements of $\hat{\mathbf{c}}$ be the maximum likelihood estimates, $\hat{c}_{st}$, as found via the framework above.  Since $\hat{\mathbf{c}}$ is asymptotically normally distributed, any linear combination of the elements is also asymptotically normally distributed.  For instance, let $a$ be a vector of the same dimension of $\hat{\mathbf{c}}$.  Then $a^t\hat{\mathbf{c}}$ is asymptotically distributed as $\mathcal{N}(a^t\mathbf{c}, a^t\Sigma a)$, where $\Sigma$ is the covariance matrix of the asymptotic distribution of $\hat{\mathbf{c}}$.

Suppose, for example, that the hypothesis $c_s$ is determined to best fit the data with $s$ ranging $s = 1, 2, 3$.  We can test to see whether or not two species are statistically equally preferred under the null hypothesis $c_{1} = c_{2}$.  This hypothesis is alternatively written in vector notation as $a^t\mathbf{c} = 0$, where $a = (1, -1, 0)^t$.  Tests of the following form $H_0: a^t\mathbf{c} = \mu$ against any alternative of interest are then approximate $Z$-tests.  Confidence intervals of any size are similarly, readily obtained.  Of course, one could also use a $t$ distribution as a small sample size correction.

### 2014-11-29 paper/simulations.tex
edit and typo check

The cluster of estimates for $c_5$ between $3.5$ and $4.0$ in the bottom panel of figure~\ref{fig:em} comes from datasets in which $X_{js5}>0$ so that $Z_{js5}=1$ for all $j,s$.  For the data shown in figure~\ref{fig:em}, this happened $73$ times out of the $500$ replicated datasets.  As mentioned above, the estimate of $c_5$ is infinite in this case. However, the EM algorithm will always provide a finite estimate for all parameters when it terminates.  In this case, we set $\tau=10^{-5}$ and it just happened that this caused the algorithm to terminate with $\hat c_5$ between $3.5$ and $4.0$ in all cases.  To confirm that this is due to the arbitrary choice of $\tau$, we repeated the algorithm with smaller values of $\tau$ for several datasets. As expected, $\hat c_5$ increased without bound as we refit the model with increasingly small values of $\tau$.

@@

The over-estimation of parameters, a symptom of the loss of information due to the unobserved counts, can also be seen with box plots of the $500$ point estimates centered at their respective true parameter values.  Figure~\ref{fig:em_bp} contains box plots of the same scenarios in Figure~\ref{fig:em}.  For the $73$ cases in which $Z_{js5} = 1$ for all $j,s$ under model $c_t$ with the huge sample size, the bias is infinite since parameter estimates will, theoretically, be infinite.  The finite bias shown in these plots is due to the finite estimates provided by the termination of the EM algorithm.  Thus, conditional on a mixture of $0$s and $1$s in the data the correspoding estimators are unbiased, but when no $0$s exist in the data the theoretical bias is infinite.

>>>>

The cluster of estimates for $c_5$ between $3.5$ and $4.0$ in the bottom panel of figure~\ref{fig:em} comes from datasets in which $X_{js5}>0$ so that $Z_{js5}=1$ for all $j,s$.  For the data shown in figure~\ref{fig:em}, this happened $73$ times out of the $500$ replicated datasets.  As mentioned above, the estimate of $c_5$ is infinite in this case. However, the EM algorithm will always provide a finite estimate for all parameters when it terminates.  In this case, we set $\tau=10^{-5}$ and it just happened that this caused the algorithm to terminate with $\hat c_5$ between $3.5$ and $4.0$.  To confirm that this is due to the arbitrary choice of $\tau$, we repeated the algorithm with smaller values of $\tau$ for several datasets. As expected, $\hat c_5$ increased without bound as we refit the model with increasingly small values of $\tau$.

@@

The over-estimation of parameters, a symptom of the loss of information due to the unobserved counts, can also be seen with box plots of the $500$ point estimates centered at their respective true parameter values.  Figure~\ref{fig:em_bp} contains box plots of the same scenarios in Figure~\ref{fig:em}.  For the $73$ cases in which $Z_{js5} = 1$ for all $j,s$ under model $c_t$ with the huge sample size, the bias is infinite since parameter estimates will, theoretically, be infinite.  The finite bias shown in these plots is due to the finite estimates provided by the termination of the EM algorithm.  Thus, conditional on a mixture of $0$s and $1$s in the data the corresponding estimators are unbiased, but when no $0$s exist in the data the theoretical bias is infinite.
